---
title: "Generalized linear regression: ZIP and ZINB"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```


# Setup

::: panel-tabset
## R

```{r, eval = F}
install.packages("pscl")
install.packages("countreg", repos="http://R-Forge.R-project.org")
install.packages("topmodels", repos="http://R-Forge.R-project.org")
install.packages("distributions3")
install.packages("modelsummary")
install.packages("marginaleffects")
install.packages("rstudioapi")
```

```{r}
library(MASS) ## glm.nb
library(countreg)
library(pscl)
library(topmodels)
library(modelsummary) 
library(boot) 
library(marginaleffects)
```

Read in the dataset for the lecture and specify column classes (types).

```{r}
df <- read.csv("../data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
head(df)
```

## Python

Load modules

```{bash, eval = F}
!pip install statsmodels 
```

```{python}
from patsy import dmatrices
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
```

Read in the dataset for the lecture and specify column classes (types).

```{python}
df = pd.read_csv("../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

## Julia

```{julia, eval = F}
using Pkg
Pkg.add(["Effects", "StatsBase", "CSV", "CategoricalArrays", "GLM"])
```

Load packages

```{julia, eval = F}
using Effects
using StatsBase
using CSV
using CategoricalArrays
using GLM
using Statistics
using DataFrames
```

Read in the dataset for the lecture and specify column classes (types).

```{julia, eval = F}
df=CSV.read("../data/polish-jvs.csv", DataFrame, 
            types = Dict(:id => Int64, :woj=> String, :public=> String, 
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```
:::

# Zero-inflated Poisson (ZIP)

::: panel-tabset
## R

```{r}
m1 <- pscl::zeroinfl(formula = vacancies ~ size + public,
                     dist = "poisson",
                     link = "logit",
                     data = df)
```

```{r}
summary(m1)
```

```{r}
rootogram(m1, plot = "base", xlim = c(0,11))
```
```{r}
avg_slopes(m1, type = "zero")
```

```{r}
avg_slopes(m1, type = "response")
```

Interpretation:

+ Count Model (Poisson component):
    + Public sector coefficient: exp(-0.580550)  ≈ 0.560 -- Being a public organization is associated with approximately 44% fewer vacancies compared to private organizations, holding other factors constant.
    + Public sector coefficient (AME):  -0.563 -- On average, public organizations have 0.563 fewer vacancies than private organizations, holding other variables constant.
    
+ Zero-Inflation Model (logit component):
    + Public sector coefficient: exp(0.02671) ≈ 1.027 -- Public organizations have approximately 2.7% higher odds of having zero vacancies compared to private organizations, but this effect is not statistically significant (p = 0.331).
    + Public sector coefficient (AME): 0.00291 -- Being a public organization is associated with a 0.291 percentage point increase in the probability of having zero vacancies compared to private organizations.
    


## Python

```{python}
y, x = dmatrices("vacancies ~ size + public", df, return_type='dataframe')
```

```{python}
m1 = sm.ZeroInflatedPoisson(endog=y, exog=x, exog_infl=x, inflation='logit').fit(maxiter = 100)
print(m1.summary())
```

## Julia
:::

# Zero-inflated Negative Binomial (ZINB)

::: panel-tabset
## R

```{r}
m2 <- pscl::zeroinfl(formula = vacancies ~ size + public,
                     dist = "negbin",
                     link = "logit",
                     data = df)
```

```{r}
summary(m2)
```

```{r}
rootogram(m2, plot = "base", xlim = c(0,11))
```

```{r}
BIC(m1, m2)
```

**Exercise**:

+ run poisson regression with the same set of variables
+ run NB regression with the same set of variables
+ which model we should use? PO, NB, ZIP or ZINB regression? Why? Justify

## Python

```{python}
m2 = sm.ZeroInflatedNegativeBinomialP(endog=y, exog=x, exog_infl=x, inflation='logit').fit(maxiter = 500)
print(m2.summary())
```

## Julia
:::

