---
title: "Generalized linear regression: Hurdle"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE, eval = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Setup

::: panel-tabset
## R

```{r, eval = F}
install.packages("geepack")
```

```{r}
library(geepack)
library(marginaleffects)
```

In this example we will use the `ohio` dataset from the `geepack`
package (source: Fitzmaurice, G.M. and Laird, N.M. (1993) A likelihood-based method for analyzing longitudinal binary responses, Biometrika 80: 141–151.)

-   The ohio data frame has 2148 rows and 4 columns. The dataset is a
    subset of the six-city study, a longitudinal study of the health
    effects of air pollution.
-   This data frame contains the following columns:
    -   `resp` -- an indicator of wheeze status (1=yes, 0=no)
    -   `id` -- a numeric vector for subject id
    -   `age` -- a numeric vector of age, 0 is 9 years old
    -   `smoke` -- an indicator of maternal smoking at the first year of
        the study

```{r}
data(ohio)
head(ohio)
```

```{r}
table(ohio$resp)
```

```{r}
unique(ohio$id) |> NROW()
```

```{r}
table(ohio$id) |> 
  table()
```

```{r}
aggregate(resp ~ smoke, data=ohio, FUN=mean)
```

**Model 1**: Simple glm model


```{r}
model1 <- glm(resp ~ age + smoke + age:smoke, family = binomial, data=ohio)
summary(model1)
```

**Model 2**: Independence (assumes no correlation between repeated measures)


```{r}
model2 <- geeglm(resp ~ age + smoke + age:smoke, 
                 id=id, 
                 data=ohio,
                 family=binomial, 
                 corstr="independence", 
                 scale.fix=TRUE)
summary(model2)
```

**Model 3**: Exchangeable (assumes constant correlation between any two time points)


```{r}
model3 <- geeglm(resp ~ age + smoke + age:smoke, 
                 id=id, 
                 data=ohio,
                 family=binomial, 
                 corstr="exchangeable", 
                 scale.fix=TRUE)
summary(model3)
```

Interpretation:

+ $\alpha = 0.355$ is the estimated within-subject correlation coefficient between any two observations from the same child. This means that any two wheeze measurements taken from the same child at different time points have a correlation of approximately 0.355.
+ About 35.5% of the variation in wheeze status can be attributed to stable, child-specific factors (genetics, chronic conditions, home environment, etc.)
+ The other ~64.5% represents time-varying factors, measurement error, and random variation

**Model 4**: AR(1) (assumes correlation decreases with time separation)

```{r}
model4 <- geeglm(resp ~ age + smoke + age:smoke, 
              id=id, 
              data=ohio,
              family=binomial, 
              corstr="ar1", 
              scale.fix=TRUE)
summary(model4)
```

Interpretation:

+  $\alpha= 0.501$ is the lag-1 autocorrelation coefficient - the correlation between consecutive time points (adjacent measurements) in the AR(1) autoregressive structure.
+ Adjacent time points (e.g., ages 9-10): correlation = $\alpha= 0.501$
+ Two time points apart (e.g., ages 9-11): correlation = $\alpha^2= 0.501^2=0.251$
+ Three time points apart (e.g., ages 9-12): correlation = $\alpha^3= 0.501^3=0.126$

```{r}
QIC(model2, model3, model4)
```

General interpretation:

+ Age effect `(-0.14 to -0.15)`: Significant negative effect (p ≈ 0.015) indicating wheeze probability decreases as children get older. 
+ Maternal smoking `(0.29 to 0.31)`: Marginally significant effect (p = 0.095-0.133) suggesting children of smoking mothers have higher wheeze rates, though this varies by model.
+ Interaction term `(0.07 to 0.08)`: Non-significant across all models (p > 0.36), indicating the age effect doesn't significantly differ between smoking and non-smoking households.

+ Models 2 and 3 tie with QIC = 1830 (independence and exchangeable)
+ Model 4 has slightly higher QIC = 1831 (AR1 is least preferred)


```{r}
avg_slopes(model2)
```


## Python

## Julia
:::

Exercise:

1. read the `beer` and limit data to single households (`size`)

```{r}
library(readxl)
beer <- read_excel("../data/beer-gee.xlsx")
beer_single <- subset(beer, size == 1)
```

2. Poisson GEE regression models:

+ formula: `beers ~ time + locality + education`
+ allow for overdispersion `scale.fix = F`
+ check different correlation structures: independent, exchangable, autocorrelated
+ which model is the best? Why?




